Leadership across cultures and contexts: an exploratory analysis

### 2020-05-08 leadership_across_cultures_contexts.Rmd
Added citation in text.
New heatmaps for features. Working on discussion.

loadd(all_data)
loadd(all_data2)
loadd(all_study_vars)
loadd(coercive_plot)
loadd(df_func)
loadd(df_qual)
loadd(features)
loadd(feature_models)
loadd(follower_benefits_txt)
loadd(follower_costs_txt)
loadd(follower_benefits_cult)
loadd(follower_costs_cult)
loadd(functions_support_cult)
loadd(functions_support_txt)
loadd(highstatus_plot)
loadd(leader_benefits_txt)
loadd(leader_benefits_cult)
loadd(leader_costs_txt)
loadd(leader_costs_cult)
loadd(m_lpca_func)
loadd(m_lpca_qual)
loadd(m_pvclust_func)
loadd(m_pvclust_qual)
loadd(multi_aic)
loadd(multi_allvars)
# loadd(p_heatmap_groups)
# loadd(p_heatmap_region)
# loadd(p_heatmap_sex)
# loadd(p_heatmap_subsis)
loadd(p_heatmap_feature_subsis)
loadd(p_heatmap_feature_region)
loadd(p_heatmap_feature_sex)
loadd(p_heatmap_feature_groups)
loadd(plot_costs_benefits)
loadd(plot_elastic_status)
loadd(plot_elastic_coercive)
loadd(plot_features_support)
loadd(plot_func_qual)
loadd(plot_pubdate)
loadd(qualities_support_cult)
loadd(qualities_support_txt)

@@

## Universality in leadership dimensions, and variation by group structure, subsistence strategy, continental region, and leader sex

@@

We first examined the `r commondim_n` leadership dimensions for which evidence was found in at least 60% of cultures. There were `r universals_n` that did not show evidence of varying by the four factors. Because evidence for these dimensions was common across cultures, the `r universals_n` unvarying dimension are candidates for universal dimensions of human leadership. See Table \@ref(tab:leaderuniversals).

@@

df_compare2 %>% 
  filter(AIC_diff >= -2 & Estimate > 0.6) %>% 
  mutate(
    Type = map_chr(var, ~ str_split(., '_')[[1]][[1]]),
    Type = str_replace(Type, '\\.', ' '),
    Type = str_to_title(Type),
    `Percent of cultures` = 100*Estimate
  ) %>% 
  select(Type, Variable, `Percent of cultures`) %>% 
  arrange(-`Percent of cultures`) %>% 
  kable(digits = 1, caption = "Candidate universal dimensions of leadership. These dimensions were common in the ethnographic record (appearing in at least 60% of cultures), and did not appear to vary by group context, subsistence strategy, continental region, or leader sex.")

@@

We then examined the `r vary_n` models (out of `r nrow(multi_allvars)` total) whose AIC values indicated improved fits over the intercept-only models, i.e., those leadership dimensions that exhibited important variation by one or more of our four factors (see Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying dimensions of leadership, one factor had a bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in about half of the dimensions. See Figure \@ref(fig:heatmapdrop1).

```{r heatmapdrop1, fig.width=8, fig.cap="Leadership dimensions that varied by group context, subsistence strategy, continental region, and/or leader sex. Colors represent the relative change in AIC value after dropping that factor from a multiple logistic mixed effects regression model. Yellow: factor improved fit relative to other factors. Black: factor did not improve fit relative to other factors. Delta AIC values were standardized by rows, and therefore can only be compared within rows."}
heatmap_drop1
```

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, for subsistence strategy, there were three leadership dimensions in which it played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)). We plotted a heatmap that displayed the relative marginal means for each subsistence strategy for each of the three leadership dimensions (see Figure S\@ref(fig:heatmapsubsis)). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastoralists had relatively low evidence for *Distribute resources*. For Group context, see Figure S\@ref(fig:heatmapgroups); for Region, see Figure, S\@ref(fig:heatmapregion); and for Leader sex, see Figure S\@ref(fig:heatmapsex).

@@

We identify these named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature. Because features were aggregated variables, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

```{r featuressupportplot, fig.width=10, fig.cap="Support of each feature variable at the levels of text records and cultures."}
plot_features_support
```

@@

We then fit separate logistic mixed effects models of each feature as a function of group context, subsistence strategy, continental region, and leader sex, with random effects for author nested within culture, comparing these to intercept-only models, and selecting all models whose AIC values were at least two points less than the intercept-only models. See Figures S\@ref(fig:heatmapfeaturesubsis), S\@ref(fig:heatmapfeatureregion), S\@ref(fig:heatmapfeaturesex), and S\@ref(fig:heatmapfeaturegroups).

@@

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. We found evidence that leaders function to resolve conflict, organization cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligence, and experienced and accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductivly, and socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). Some common leadership dimensions also did not vary by group context, continental region, subsistence strategy, or leader sex. These included qualities, such as being knowledgeable and intelligence, wealthy, strategically nepotistic, and generous, and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal", we do not mean that all leaders have all these traits. Instead, we mean that there is a fairly high level of evidence for these traits across cultures that does not vary by continental region, subsistence strategy, group context, or leader sex. See Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport).

@@

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories of leadership which have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigous leaders [but see @garfield_investigating_2019]. These dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015], but that model does not emphasize the concrete material benefits seen here. Regarding the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity and status leveling mechanisms, the absence of *High status* and *Aggressiveness* on this universal list is notable. Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019].

The following quote, from the Pawnee [@murrayXX], illustrates leadership involving intelligence, conflict resolution, and generosity:

@@

Evidence for most dimensions of leadership was absent in most cultures: of our `r length(all_study_vars)` dimensions of leadership, there was evidence in less than 50% of cultures for `r sum(df_compare2$Estimate < 0.5)` of them. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb).

@@

## Relative effects of subsistence strategy, region, leader sex, and group structure on all variables

@@

```{r heatmapfeaturesubsis, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each feature variable by subsistence strategy. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}
plot(p_heatmap_feature_subsis)
```

```{r heatmapfeatureregion, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each feature variable by continental region. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}
plot(p_heatmap_feature_region)
```

```{r heatmapfeaturesex, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each feature variable by leader sex. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}
plot(p_heatmap_feature_sex)
```

```{r heatmapfeaturegroups, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each feature variable by group context. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}
plot(p_heatmap_feature_groups)
```

>>>>

loadd(all_data)
loadd(all_data2)
loadd(all_study_vars)
loadd(coercive_plot)
loadd(df_func)
loadd(df_qual)
loadd(features)
loadd(feature_models)
loadd(feature_models_aic)
loadd(follower_benefits_txt)
loadd(follower_costs_txt)
loadd(follower_benefits_cult)
loadd(follower_costs_cult)
loadd(functions_support_cult)
loadd(functions_support_txt)
loadd(highstatus_plot)
loadd(leader_benefits_txt)
loadd(leader_benefits_cult)
loadd(leader_costs_txt)
loadd(leader_costs_cult)
loadd(m_lpca_func)
loadd(m_lpca_qual)
loadd(m_pvclust_func)
loadd(m_pvclust_qual)
loadd(multi_aic)
loadd(multi_allvars)
# loadd(p_heatmap_groups)
# loadd(p_heatmap_region)
# loadd(p_heatmap_sex)
# loadd(p_heatmap_subsis)
loadd(p_heatmap_feature_subsis)
loadd(p_heatmap_feature_region)
loadd(p_heatmap_feature_sex)
loadd(p_heatmap_feature_groups)
loadd(plot_costs_benefits)
loadd(plot_elastic_status)
loadd(plot_elastic_coercive)
loadd(plot_features_support)
loadd(plot_func_qual)
loadd(plot_pubdate)
loadd(qualities_support_cult)
loadd(qualities_support_txt)

@@

## Universality in leadership dimensions

@@

We first examined the `r commondim_n` leadership dimensions for which evidence was found in at least 60% of cultures. Of these, `r universals_n` did not meaningfully vary by the four factors. Because evidence for these dimensions was common across cultures, the `r universals_n` unvarying dimension are candidates for universal dimensions of human leadership. See Table \@ref(tab:leaderuniversals).

@@

df_compare2 %>% 
  filter(AIC_diff >= -2 & Estimate > 0.6) %>% 
  mutate(
    Type = map_chr(var, ~ str_split(., '_')[[1]][[1]]),
    Type = str_replace(Type, '\\.', ' '),
    Type = str_to_title(Type),
    `Percent of cultures` = 100*Estimate
  ) %>% 
  select(Type, Variable, `Percent of cultures`) %>% 
  arrange(-`Percent of cultures`) %>% 
  kable(digits = 1, caption = "Candidate universal dimensions of leadership. These dimensions were common in the ethnographic record (appearing in at least 60% of cultures), and did not appear to meaningfully vary by group context, subsistence strategy, continental region, or leader sex.")

@@

## Variation by group structure, subsistence strategy, continental region, and leader sex

We then examined the `r vary_n` (out of `r nrow(multi_allvars)` total) leadership dimensions that exhibited important variation by one or more of our four factors, i.e., whose AIC values indicated improved fits over the intercept-only models (see Tables S\@ref(tab:variablevars) and S\@ref(tab:universalvars)). To determine which of the four factors contributed to improved fit, we dropped each factor in turn and computed the change in AIC. For most of these `r vary_n` varying dimensions of leadership, one factor had a noticably bigger effect on fit than the others. Variation in group context, in particular, was associated with variation in almost half of the dimensions. See Figure \@ref(fig:heatmapdrop1)A.

```{r heatmapdrop1, fig.width=12, fig.cap="Leadership dimensions and features that varied by group context, subsistence strategy, continental region, and/or leader sex. A. Leadership dimensions. B. Leadership features. Colors represent the relative change in AIC value after dropping that factor from a multiple logistic mixed effects regression model. Yellow: including factor improved fit relative to other factors. Black: including factor did not improve fit relative to other factors. Delta AIC values were standardized by rows, and therefore can only be compared within rows."}
heatmap_drop1 + guides(fill='none') + heatmap_features_drop1 + plot_annotation(tag_levels = 'A')
```

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, for subsistence strategy, there were three leadership dimensions in which it played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For Group context, see Figure S\@ref(fig:heatmapgroups); for Region, see Figure, S\@ref(fig:heatmapregion); and for Leader sex, see Figure S\@ref(fig:heatmapsex).

@@

We identify these seven named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to variables that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all variables within these clusters were clearly related to the themes of each feature. Because features were aggregated variables, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

```{r featuressupportplot, fig.width=10, fig.cap="Support of each feature variable at the levels of text records and cultures. Feature support at the text record was defined as the adjusted proportion of text records with evidence for at least one underlying variable in the feature. Feature support at the culture level was defined as the proportion of cultures with evidence for the feature in at least text record in the culture."}
plot_features_support
```

@@

We then fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, with random effects for author nested within culture, comparing these to intercept-only models, and selecting the `r nrow(feature_models_aic)` models whose AIC values were at least two points less than the intercept-only models (only Cultural conformity was dropped). As with our leader dimension models, we then dropped each factor variable in turn and compared the change in AIC. Large increases indicated that including the factor substantially improved fit. See Figure \@ref(fig:heatmapdrop1)A. We then compared the levels of each factor that improved fit for each feature variable. See Figures S\@ref(fig:heatmapfeaturesex), and S\@ref(fig:heatmapfeaturegroups).

@@

We found strong evidence that several dimensions of leadership were universal across cultures, and across group contexts within cultures. Leaders function to resolve conflict, organize cooperation, and provide counsel and direction in over 70% of cultures, and are seen as high status, knowledgeable and intelligent, and experienced and accomplished in about 80% or more of cultures (see Figure \@ref(fig:varsupport)). Leaders benefited materially, reproductivly, and socially in 50% or more of cultures (see Figure \@ref(fig:varsupportcb)). Some common leadership dimensions also did not vary by the four factors we examined: group context, continental region, subsistence strategy, and leader sex. These dimensions included qualities, such as being knowledgeable and intelligent, wealthy, strategically nepotistic, and generous, and functions, such as resolving conflict and representing the group. These dimensions are therefore candidates for universal dimensions of human leadership. By "universal" we do not mean that all leaders have all these traits. Instead, we mean that there is evidence for these traits across a solid majority of cultures that does not vary much by the four factors. See Table \@ref(tab:leaderuniversals) and Figure \@ref(fig:varsupport).

@@

Leaders can be conceptualized as performing three broad functions: providing goods and services (Prosocial), organizing collective actions, and

The leadership "features" we derived by clustering our leadership dimensions were represented across most cultures. See Figure \@ref(fig:featuressupportplot).

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories of leadership that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigous leaders [but see @garfield_investigating_2019]. These dimensions also partially correspond to the service-for-prestige model [@price_evolution_2014; @price_service-for-prestige_2015], but that model does not emphasize the concrete material benefits seen here. The absence of *High status* and *Aggressiveness* on this universal list corresponds to the reverse dominance hierarchy model [@boehm_egalitarian_1993], which emphasizes knowledge, intelligence, conflict resolution, generosity and status leveling mechanisms.

```{r matingbenefit}
matingbenefit <- signif(100*leader_benefits_cult$Estimate[leader_benefits_cult$Variable=='Mating benefit'], 3)
```

Most potentially universal dimensions discovered here appear in a model developed by @hagen_leadership_2019, in which knowledgable and intelligent individuals provide computational (cognitive) services to group members in exchange for a variety of fitness benefits, and these services often involve conflict resolution and other social functions [see also @garfield_evolutionary_2019]. But the computational services model emphasizes mating benefits, which appear in `r matingbenefit`% of cultures, which was below our arbitrary 60% threshold for "universality". See Figure \@ref(fig:varsupportcb).

Pawnee chiefs illustrate leadership involving several of the universal dimesions, including intelligence, conflict resolution, and generosity [@murrayXX]:

@@

There was substantial cross-cultural variation in evidence for most dimensions of leadership: of our `r length(all_study_vars)` dimensions of leadership, `r sum(df_compare2$Estimate < 0.5)` had evidence in less than 50% of our `r nrow(leader_cult)` cultures. See Figures \@ref(fig:varsupport) and \@ref(fig:varsupportcb). There was *systematic* variation in `r vary_n` leadership dimensions by continental region, subsistence, strategy, group context, and/or leader sex, factors for which we had complete data for all text records. Interestingly, these systematically varying dimensions tended to be associated with one factor above the others. See Figure \@ref(fig:heatmapdrop1). The clearest pattern to emerge from these analyses is that, not too surprisingly, many dimensions of leadership are context-specfic: Aggressiveness and Bravery were associated with military groups, Supernatural and Ritual functions were associated with religious groups, Organizing cooperation and Experience/accomplished were associated with economic groups, and Providing counsel and older Age were associated with kin groups. See Figure \@ref(fig:heatmapgroups).

@@

## Relative effects of subsistence strategy, region, leader sex, and group structure on leadership dimension

@@

```{r heatmapfeaturesubsis, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each feature variable by subsistence strategy. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}
# plot(p_heatmap_feature_subsis)
```

```{r heatmapfeatureregion, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each feature variable by continental region. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}
# plot(p_heatmap_feature_region)
```

```{r heatmapfeaturesex, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each feature variable by leader sex. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}
heatmap_feature_sex
```

```{r heatmapfeaturegroups, fig.cap="Heatmap of relative estimated marginal means of the levels of evidence for each feature variable by group context. Estimated marginal means were standardized by rows. The colors therefore represent the relative amount of evidence within the same outcome variable."}
heatmap_feature_groups
```
